---
title: "Exploratory Data Analysis (EDA) of 2022 High Volume For-Hire Vehicles"
format: 
  gfm:
    toc: true

execute:
  message: false
  warning: false
---

After completing the [business understanding](https://github.com/AngelFelizR/nyc-taxi-project/tree/master/notebooks/02-business-understanding) step, we have a clear objective in mind and an initial description for each column the [raw data](https://github.com/AngelFelizR/nyc-taxi-project/tree/master/data), we are ready to perform the *data understanding* by performing an EDA with the following steps:

1. Examining the distribution of each individual variable by counting the categorical variables and creating histograms or box plots for numerical variables
2. Defining the target variable and confirming its distribution
3. Confirming domain knowledge relations by creating visualization with 2 or more variables
4. Taking a subset of the data to fit in RAM
5. Exploring correlations between predictors by using a correlation matrix or running a PCA
6. Removing high correlated predictors
7. Exploring correlations between the target and predictors creating a correlation funnel and some scatter plots.

After completing this process, we will have the following outcomes:

- Confirming the meaning of each variable
- Ensuring data quality by finding missing values and 
- Identifying the best models to train
- Creating new features that can enhance the predictive power of the machine learning model


## Setting the environment up

To setting the `R` environment up we just need to apply the following 4 steps:

1. Loading the packages to use.

```{r}
library(here)
library(data.table)
library(ggplot2)
library(scales)
library(forcats)
library(dplyr)
library(arrow)
```

2. Sourcing the following custom functions to avoid repeating myself:
  - `count_pct`: It counts the number rows where each unique value repeated in the columns selected arranging there results in descent order and adds a percentage column after collecting the results from `arrow`.
  - `join_zones`: It creates the `start_borough`, `start_zone` and `start_service_zone` columns based on the `PULocationID` column and the `end_borough`, `end_zone` and `end_service_zone` columns based on the `DOLocationID` column.

```{r}
source(here("R/01-custom-functions.R"))
```

3. Creating an Arrow connection object to perform some manipulations in disk before taking the data into the RAM memory.

```{r}
NycTrips2022 <- 
  here("data/trip-data/year=2022") |>
  open_dataset() |>
  mutate(company = case_when(
    hvfhs_license_num == "HV0002" ~ "Juno",
    hvfhs_license_num == "HV0003" ~ "Uber",
    hvfhs_license_num == "HV0004" ~ "Via",
    hvfhs_license_num == "HV0005" ~ "Lyft"
  )) |>
  select(-hvfhs_license_num)
```

4. Importing the zone code description.

```{r}
ZoneCodes <- fread(
  here("data/taxi_zone_lookup.csv"),
  colClasses = c("integer", "character", "character", "character")
)
```

5. Counting the number of trips for each code, collecting and translating the zone codes.

```{r}
TripsZoneDistribution <-
  NycTrips2022 |>
  count(PULocationID, DOLocationID) |>
  collect() |>
  join_zones(zone_tb = ZoneCodes)
```


## Exploring distribution of each individual variable

### Categorical variables

- `company`: The majority number of trips are done by *Uber* (HV003) and the rest for *Lyft*.

```{r}
NycTrips2022 |> count_pct(company)
```

- `dispatching_base_num`: For most of the trips the dispatching number is **B03404** for Uber and **B03406** related for Lyft. Uber also has other 27 dispatching number, but they are really weird to see.

```{r}
NycTrips2022 |>
  count_pct(company, dispatching_base_num) |>
  as.data.table() |>
  (\(dt) dt[, .(n_unique_code = n_distinct(dispatching_base_num),
                n_missing = sum(is.na(dispatching_base_num)),
                n = sum(n),
                pct = sum(pct) |> round(4L)),
            by = .(company,
                   common = pct > 0.01)
          ][order(-n)] )()
```

- `originating_base_num`: This column has the same information as `dispatching_base_num` or has missing values, so we don't need to use it.

```{r}
NycTrips2022 |>
  count_pct(company,
            same_id = originating_base_num == dispatching_base_num)
```

- `shared_request_flag`: Most of passengers don't agree to a shared/pooled ride.

```{r}
NycTrips2022 |> count_pct(shared_request_flag)
```

- `shared_match_flag`: From the subset of trips that agreed to share the ride only 24% achieved the match, we even can see some weird cases when the passenger did not initially agree to a shared/pooled ride (shared_request_flag = N), but ended up sharing the vehicle with another passenger who booked separately at some point during the trip (shared_match_flag = Y).

```{r}
NycTrips2022 |> 
  count(shared_request_flag, shared_match_flag) |>
  collect() |>
  group_by(shared_request_flag) |>
  mutate(pct = n / sum(n)) |>
  arrange(desc(shared_request_flag), desc(shared_match_flag))
```

- `access_a_ride_flag`: *Uber* isn't reporting whether their trips were administered on behalf of the Metropolitan Transportation Authority and for *Lyft* the answer is always "N", so we will **erase** this column as it doesn't show any useful information.

```{r}
NycTrips2022 |> count_pct(company, access_a_ride_flag)
```

- `wav_request_flag`: It's really unusual for a passenger to request a wheelchair-accessible vehicle.

```{r}
NycTrips2022 |> count_pct(wav_request_flag)
```

- `wav_match_flag`: If someone ask for a wheelchair-accessible vehicle it's really often that passenger get that type of vehicle.

```{r}
NycTrips2022 |> 
  count(wav_request_flag, wav_match_flag) |>
  collect() |>
  group_by(wav_request_flag) |>
  mutate(pct = n / sum(n)) |>
  arrange(desc(wav_request_flag), desc(wav_match_flag))
```

- `start_borough` and `end_borough`: After exploring the next distribution of trips we can highlight that:

  - 94% of the trips take place in *Manhattan*, *Brooklyn*, *Queens* and *Bronx*. As consequence, there is no reason to keep exploring trips related to *Staten Island*, *EWR* or *Unknown* as they are the minority.
  - 70% the trips take place without going to outside the start borough.

```{r}
TripsZoneDistribution[, .(n = sum(n)),
                      by = c("start_borough", "end_borough")
][order(n)
][, c("start_borough", "end_borough") := 
    lapply(.SD, \(x) factor(x, levels = unique(x, fromLast = TRUE)) ),
  .SDcols = c("start_borough", "end_borough")
][, end_borough := fct_rev(end_borough)] |>
  ggplot(aes(end_borough, start_borough))+
  geom_tile(aes(fill = n)) +
  geom_text(aes(label = percent(n/sum(n), accuracy = 0.01))) +
  scale_fill_gradient(low = "white", 
                      high = "red",
                      labels= comma_format())+
  scale_x_discrete(position = "top") +
  labs(title = "Distribution of Trips by Borough in NYC 2022",
       x = "Trip End", 
       y = "Trip Start", 
       fill = "Number of Trips") +
  theme_classic() +
  theme(plot.title = element_text(face = "bold"),
        axis.ticks = element_blank(),
        axis.line = element_blank(),
        axis.text = element_text(color = "black"),
        axis.title = element_text(face = "italic"))
```

- `start_service_zone` and `end_service_zone`: 52% of the trips take place at the *Boro Zone*, 26% of trips take place at the *Yellow Zone* and only small fraction goes to the *Airports*.

```{r}
TripsZoneDistribution[  
  !start_borough %chin% c("Staten Island", "Unknown", "EWR") &
    !end_borough %chin% c("Staten Island", "Unknown", "EWR"), 
  .(n = sum(n)),
  by = c("start_service_zone",
         "end_service_zone")
][order(n)
][, c("start_service_zone", "end_service_zone") := 
    lapply(.SD, \(x) factor(x, levels = unique(x, fromLast = TRUE)) ),
  .SDcols = c("start_service_zone", "end_service_zone")
][, end_service_zone := fct_rev(end_service_zone)] |>
  ggplot(aes(end_service_zone, start_service_zone))+
  geom_tile(aes(fill = n)) +
  geom_text(aes(label = percent(n/sum(n), accuracy = 0.01))) +
  scale_fill_gradient(low = "white", 
                      high = "red",
                      labels= comma_format())+
  scale_x_discrete(position = "top") +
  labs(title = "Distribution of Trips by Service Zone in NYC 2022",
       x = "Trip End", 
       y = "Trip Start", 
       fill = "Number of Trips") +
  theme_classic() +
  theme(plot.title = element_text(face = "bold"),
        axis.ticks = element_blank(),
        axis.line = element_blank(),
        axis.text = element_text(color = "black"),
        axis.title = element_text(face = "italic"))
```


```{r}
#| echo: false
#| include: false

ZoneUniqueComb <-
  TripsZoneDistribution[ 
    !start_borough %chin% c("Staten Island", "Unknown", "EWR") & 
      !end_borough %chin% c("Staten Island", "Unknown", "EWR"),
    unique(.SD),
    .SDcols = c("start_zone", "end_zone")
  ][, .N] |>
  comma()

EndingUniqueZones <-
  TripsZoneDistribution[ 
    !start_borough %chin% c("Staten Island", "Unknown", "EWR") & 
      !end_borough %chin% c("Staten Island", "Unknown", "EWR"),
    uniqueN(end_zone),] |>
  comma()
```


- `start_zone` and `end_zone`: As our data has `r ZoneUniqueComb` rows of relations between both columns, we opted to transform data in a way what each unique zone represent a row reducing the points to be plotted to only `r EndingUniqueZones` by following the next steps: 

  1. Summarizing the total number of trips for each **starting point** independently to its destination 
  2. Summarizing the total number of trips for each **ending point** independently to its origin.
  3. Joining both tables into one.

```{r}
# 1. Summarizing Staring Zones
StartingZonesCount <-
  TripsZoneDistribution[
    !start_borough %chin% c("Staten Island", "Unknown", "EWR") &
      !end_borough %chin% c("Staten Island", "Unknown", "EWR"),
    .(start_trips = sum(n)),
    by =  .(borough = start_borough, 
            zone = start_zone)
  ]

# 2. Summarizing Ending Zones
EndingZonesCount <-
  TripsZoneDistribution[
    !start_borough %chin% c("Staten Island", "Unknown", "EWR") &
      !end_borough %chin% c("Staten Island", "Unknown", "EWR"), 
    .(end_trips = sum(n)),
    by =  .(borough = end_borough, 
            zone = end_zone)
  ]

# 3. Inner Joining Starting and Ending Zones Counts
JoinedZonesCount <-
  StartingZonesCount[
    EndingZonesCount,
    on = c("borough", "zone"),
    nomatch = 0
  ]
```

Once we have a much simpler data to work with, we is easy to confirm with the next linear regression that `start_trips` and `end_trips` has almost the same values the model has an slope of one. That means that in must of cases **if someone takes a taxi to go to any place it's really likely to take another taxi back**.

```{r}
lm(end_trips ~ start_trips, 
   data = JoinedZonesCount) |>
  summary()
```

Let's now explore the zones where there is **no balance** between the `start_trips` and the `end_trips` in the most visited zoned of each Borough. To do so, we defined the rate `end_trips`/`start_trips` and highlight zones with lower rate than the 15% percentile or higher rate than 85% percentile.

```{r}
# Creating dataset to plot
ZonesCountToPlot <-
  copy(JoinedZonesCount)[
    j = `:=`(ending_starting_rate = end_trips/start_trips,
             borough = fct_reorder(borough, -end_trips, .fun = sum, na.rm = TRUE),
             end_m_trips = end_trips / 1e6L,
             start_m_trips = start_trips / 1e6L)
  ][, unbalance_situation := fcase(
    ending_starting_rate < quantile(ending_starting_rate, 0.15),
    "More starts than ends",
    ending_starting_rate > quantile(ending_starting_rate, 0.85),
    "More ends than starts",
    default = "Balanced"
  )
  ][order(-(start_trips + end_trips)), 
    .SD[1:6],
    by = "borough"] 

# Creating the scatted plot
ggplot(ZonesCountToPlot,
       aes(start_m_trips, end_m_trips))+
  geom_blank(aes(pmax(start_m_trips, end_m_trips),
                 pmax(start_m_trips, end_m_trips)))+
  geom_abline(linewidth = 0.8,
              alpha = 0.5)+
  geom_point(aes(color = borough),
             size = 3.5,
             alpha = 0.75)+
  geom_text(data = ZonesCountToPlot[unbalance_situation ==
                                      "More starts than ends"],
            aes(label = zone),
            size = 3.5,
            hjust = -0.12,
            check_overlap = TRUE)+
  geom_text(data = ZonesCountToPlot[unbalance_situation ==
                                      "More ends than starts"],
            aes(label = zone),
            size = 3.5,
            hjust = 1.12,
            check_overlap = TRUE)+
  scale_x_continuous(labels = comma_format(accuracy = 0.1, suffix = " M"))+
  scale_y_continuous(labels = comma_format(accuracy = 0.1, suffix = " M"))+
  coord_equal() +
  labs(title = "Top 6 Most Important Zones by Borough",
       color = "Borough",
       x = "Number of Trips Starting",
       y = "Number of Trips Ending")+
  theme_light()+
  theme(legend.position = "top",
        text = element_text(color = "black"),
        plot.title = element_text(face = "bold"))
```

Based on the results, we can highlight the next points:

1. The airports present in Queens, *LaGuardia Airport* and *JFK Airport*, have many more trips going to the airport than going out of airport. This might happen due that there are more transportation options like other taxis, shuttles, and public transportation.

2. The remaining zones, *Jackson Heights*, *East Village* and *TriBeCa/Civic Center*, are residential zones with a variety of public transportation options.

### Time variables

- `month`:
- `request_datetime`:
- `on_scene_datetime`:
- `pickup_datetime`:
- `dropoff_datetime`:


### Numerical variables

- `trip_miles`:
- `trip_time`:
- `base_passenger_fare`:
- `tolls`:
- `bcf`:
- `sales_tax`:
- `congestion_surcharge`:
- `airport_fee`:
- `tips`:
- `driver_pay`:


