---
title: "NYC Trip - Exploratory Data Analysis"
format: gfm
---

Even thought the main purpose of this project is to predict the tips that taxi drivers received from passenger is important to have general understanding of the data to **avoid ending with wrong conclusions**.

To perform this EDA I will:

- **Understanding the distribution of each variable:** That will give a reference of the normality.

- Validate if the data has missing values
- Create new features base on domain knowledge
- Confirm assumptions

## Setting-up the enviroment

1. Loading main packages

```{r}
library(data.table)
library(lubridate)
library(timeDate)
library(ggplot2)
library(scales)
library(patchwork)
theme_set(theme_light())

source("01-scripts/02-custom-functions.R")
```

2. Importing from data from January to March as our training data with **56,853,541 rows** as our training corresponding to **10 Gb**.

```{r}
TripDataTrain <- fst::read_fst(
  path = "00-data/TripDataTrain.fst",
  as.data.table = TRUE
)


TripDataTrain[, report_month := NULL]

TripDataDim <- dim(TripDataTrain)
comma(TripDataDim)
object.size(TripDataTrain) |> format(unit = "auto")
```

3. Down sampling the data to **17,000,000 rows** to solve memory limitations problems, but keeping a representative sample of the population.

```{r}
set.seed(20230922)
TripDataTrain <- TripDataTrain[sample.int(TripDataDim[1L],17e6)]

TripDataDim <- dim(TripDataTrain)
comma(TripDataDim)
object.size(TripDataTrain) |> format(unit = "auto")
```

3. Decoding data based on dictionary information.

```{r}
TripDataTrain <- decode_cols(
  trip_table = TripDataTrain,
  zone_path = "00-data/taxi+_zone_lookup.csv"
)
```

## Validating the distribution of each variable

### Categorical variables

By counting the rows by category for each row we found that:

- We can **remove** the `dispatching_base_num` and `originating_base_num` as theirs information is really close to the `hvfhs_license_num` and their other values only represent **0.02%** of the rows.
- Based on `shared_request_flag` and `shared_match_flag` we know that must of the passengers **agree to shared the ride** even if other passenger booked separately.
- Based on `access_a_ride_flag` we know that must of the **trips were administered by MTA**. We also could that the field is missing the "Y" flag.
- Based on `wav_request_flag` and `wav_match_flag` we know that only **0.16%** of trips requested a wheelchair-accessible vehicle but **6.84%** of trips had that capacity with are really good news.
- In the `PU_Borough` and `DO_Borough`columns very few trips pass through "Staten Island", "Unknown" or "EWR" so we will consolidate them as "Other".
- `PU_Zone` and `DO_Zone` have too many categories and we don't need them for this process.
- As "EWR" represent the **Newark Liberty International Airport** and represent very few trips, we can consolidate this category to the must general one "Airports" for the `PU_service_zone` and `DO_service_zone` columns

```{r}
cat_vars <- TripDataTrain[, names(.SD), .SDcols = is.character]

for(cat_i in cat_vars){
  TripDataTrain[, .(count = .N), 
                by = cat_i
  ][, pct_count := round(count/sum(count)*100, 2)
  ][order(-count)] |>
  print()
}

```

### Numerical variables

To validate numeric variables we are potting at normal and logarithmic scale.

```{r}
num_vars <- TripDataTrain[, names(.SD), .SDcols = is.numeric]

for(num_i in num_vars){
  
 print(custom_histogram(TripDataTrain, num_i))

}

```


### Date variables

To validate numeric variables we are potting at normal and logarithmic scale.

```{r}
date_vars <- TripDataTrain[, names(.SD), .SDcols = is.POSIXct]

for(date_i in date_vars){
  
  date_hist <-
    ggplot(TripDataTrain, aes(get(date_i)))+
    geom_histogram(fill = "blue",
                   color = "black",
                   alpha = 0.8,
                   bins = 30)+
    labs(title = paste0(date_i, " Distribution"),
         x = date_i)
  
  print(date_hist)
}

```


## Missig values validation

```{r}
ggplot(TripDataTrain, aes(trip_miles, trip_time))+
  geom_point(alpha = 0.25)

TripDataTrain[trip_miles <= 0, 
              ..num_vars
][, id := .I
][, melt(.SD,
         id.vars = "id",
         variable.factor = FALSE)
][, .(min_value = min(value),
      max_value = max(value)),
  by = "variable"]

TripDataTrain[trip_time <= 0, 
              ..num_vars
][, id := .I
][, melt(.SD,
         id.vars = "id",
         variable.factor = FALSE)
][, .(min_value = min(value),
      max_value = max(value)),
  by = "variable"]

TripDataTrain[driver_pay <= 0, 
              ..num_vars
][, id := .I
][, melt(.SD,
         id.vars = "id",
         variable.factor = FALSE)
][, .(min_value = min(value),
      max_value = max(value)),
  by = "variable"]

TripDataTrain[driver_pay < 0, View(.SD)]

```


Only `originating_base_num` and `on_scene_datetime`

```{r}
TripDataTrain[, lapply(.SD, \(x) sum(is.na(x)))
][, id := 1L
][, melt(.SD,
         id.vars = "id",
         variable.factor = FALSE,
         value.name = "n_miss")
][order(-n_miss), 
  !c("id")
][n_miss>0
][, pct_miss := n_miss/nrow(TripDataTrain)*100][]
```


## Understanding the distribution of each variable

1. **hvfhs_license_num:** Must of the trips are Uber related.

```{r}
TripDataTrain[, .N, 
              by = .(hvfhs_license_num = TaxiCompanyName[hvfhs_license_num])] |>
  ggplot(aes(N, hvfhs_license_num))+
  geom_col(aes(fill = hvfhs_license_num),
           width = 0.65)+
  scale_x_continuous(labels = comma_format(accuracy = 1))+
  scale_fill_brewer(palette = "Dark2")+
  labs(title = "Number of Trips by Company",
       y = "Company",
       x = "Trip Count")+
  theme(plot.title = element_text(face = "bold", hjust = 0.5),
        legend.position = "none",
        panel.grid.major.y = element_blank())
```


2. **dispatching_base_num:** Doesn't provide much information in relation in relation to the `hvfhs_license_num`.

```{r}
TripDataTrain[, .N, 
              by = .(company = TaxiCompanyName[hvfhs_license_num],
                     dispatching_base_num)
][order(company,-N)
][, prop_N := percent(N/sum(N), accuracy = 0.01),
  by = "company"][]
```

3. **originating_base_num:** Doesn't provide much information in relation to `dispatching_base_num`.

```{r}
TripDataTrain[, .N, 
              by = .(same_base = dispatching_base_num == originating_base_num,
                     missing_dispatching_base_num = is.na(dispatching_base_num),
                     missing_originating_base_num = is.na(originating_base_num))
][, prop_N := percent(N/sum(N), accuracy = 0.01)][]

```


```{r}
TripDataTrain[, .N, 
              by = .(company = TaxiCompanyName[hvfhs_license_num],
                     originating_base_num)
  ][order(company,-N)
  ][, prop_N := percent(N/sum(N), accuracy = 0.01),
    by = "company"][]
```


3. **Pickup_datetime by month:**

```{r}
TripDataTrain[, .N,
              by = .(month = floor_date(request_datetime, unit = "day"))] |>
  ggplot(aes(month, N))+
  geom_line()
```



## Creating new features

```{r}

```


